---
title: "In-Class Exercise 1: Now You See It!"

date: "January 13, 2024"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
---

## Loading R Packages

In this hands-on exercise, two R packages will be used:

-   [tidyverse](https://www.tidyverse.org/)

-   [haven](https://haven.tidyverse.org/)

The code chunk used is as follows:

```{r}
pacman::p_load(tidyverse, haven)
```

## Importing PISA Data

The code chunk below uses [`read_sas()`](https://haven.tidyverse.org/reference/read_sas.html) function of [`haven`](https://haven.tidyverse.org/) to import PISA data into the R environment.

`#| eval: false` was used to prevent the code chunk from running.

```{r}
#| eval: false
stu_qqq <- read_sas("Data/cy08msp_stu_qqq.sas7bdat")
```

The code chunk below uses the [`filter()`](https://dplyr.tidyverse.org/reference/filter.html) function of [`dplyr`](https://dplyr.tidyverse.org/) to retrieve records from Singapore (SGP) only.

```{r}
#| eval: false
stu_qqq_SG <- stu_qqq %>% 
  filter(CNT == "SGP")
```

The code chunks below uses `write_rds()` and `read_rds()` to save and read the records from Singapore in R format.
```{r}
#| eval: false
write_rds(stu_qqq_SG, 
          "Data/stu_qqq_SG.rds")
```

```{r}
stu_qqq_SG <- read_rds("data/stu_qqq_SG.rds")
```
